Can a High Color Contrast Touch Interface Increase User Reaction Time when Using a Smart Phone Web Based Application?

### 2016-01-03 Paper_revised.tex
revision

\title{Can a High Color Contrast Touch Interface Increase User Reaction Time when Using a Smart Phone Web Based Application.}

@@

%%%%% BEGIN OF 2nd ADJUSTMENTS SECTION %%%%%
% Do NOT include an abstract in the outline and annotated bibliography
% (deliverable 2). For the full and final paper (deliverables 3 and 4a/b),
% you must uncomment the following two commands and place your text for
% the abstract in between them.
%
\begin{abstract}
This study aims to investigate if it is possible to receive better user performance when using a high color contrast compared to a low color contrast touch screen interface in a mobile web application. The performance was measured by combining the time each user took to respond to instructions (reaction time) and the number of input errors the user made. Received test data appeared to be corrupt, most probably due to limitations in web application technology, despite this our result show that the low color contrast interface produced 475\% more user input errors compared to the high color contrast interface. No difference in reaction time could be found between the two versions.
\end{abstract}
%
%
%This study aims to investigate if there is a performance difference between a high color contrast and a low color contrast touch screen interface. The performance was measured by combining the time each test subject took to respond on instructions (reaction time) and the number of input errors. The result shows that the high color contrast interface resulted in significantly fewer user input errors. There was no differences in reaction time between the two interfaces.
%%%%% END OF 2nd ADJUSTMENTS SECTION %%%%%

\section{Introduction}
The main goal with this paper is to evaluate if color contrast has a significant impact on usability in web based smart-phone applications. Consumer touch screen devices such as smart-phones have rapidly increased in amount and availability recent years. The touch screen technology have made great advances~\cite{jennings2013touch} and it is frequently used as a way of receiving user input. When this technology moves to a broader audience, higher demands on usability needs to be set~\cite{gong2004guidelines}. To further explore how usability can be improved using touch screen technology we have in this paper investigated if reaction time (the time it takes for a user to react on instructions, further referred to as RT in this paper) and user input errors (the number of user input errors when interacting with a system, further referred to as UIE in this paper) can reach better results by increasing color contrast within the interface. Better results are assumed as faster RTs and less number of UIEs.

Earlier studies completed in this field have explored how we perceive different color contrast and how it can affect our reading performance~\cite{wu2003improving}. It has also been proved that chromaticity, contrast, and cone opponency in color space can affect RTs~\cite{mckeefry2003simple}. Many best practices for mobile development have also shown that "requirements for sufficient color contrast"~\cite{marcus2013design} must be met to best exaggerate the content. Still the question remains, can a high color contrast interface increase the the reaction time and decrease the amount of user input errors?

High RTs and low UIEs are especially desirable when designing user interfaces for situations with high demands on quick user input and low error tolerance such as in emergency situations. The results of this paper can be used by any designer or developer.

\section{Method}
In order to be able to test if color contrast has a substantial impact on the RT and UIE of a touch screen user interface we have designed a simple web application for an Android smart-phone. The application exists in two versions, one with low color contrast (referred to as LCC in this paper) and one with high color contrast (referred to as HCC in this paper) interface as can be seen in Fig~\ref{fig:application}. We conducted an A/B test to measure the performance between the versions.

@@

We designed the application by following design guidelines given here~\cite{hoober2011designing,johnson2013designing,gong2004guidelines,norman2013design}. These guidelines can be seen as a set of rules to follow when designing any kind of user interface. We used guidelines to get proper sized buttons, good spacing between content and best positioning of navigation.

@@

The application consists of 4 buttons\footnote{On screen button}, each button representing a function. To differentiate each button from the others we put a unique shape in each button. The shapes we used were a, square, circle, triangle and a rhomb (Fig~\ref{fig:application}).

To be able to give the test persons consistent instructions through the complete test we designed a program that presented the instructions on a secondary screen. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{instructions}
	\caption{Our program designed to present user instructions. Instructions were given in a series of 20. Between each instruction there was a delay set to be between 2.5 and 4 seconds. Each time a new symbol where showed a timestamp were logged in a unique user instruction file.
	\label{fig:instructions}}
\end{figure}
The instructions were a series of 20 instructions with shapes identical to the ones in the application buttons. The test persons were told to press the representative shape in the application as fast as they could in order to measure the reaction time. If they pressed the wrong button they were told to continue the test by pressing the right button, the application registered this as an UIE. Each test was performed indoors with varying surroundings such as different lightning conditions and noise levels.

@@

\subsection{Evaluation of the A/B test}
Our data received from the tests consisted of UNIX timestamps\footnote{More info at: \url{http://www.unixtimestamp.com/}} measured in milliseconds. Each time the user interacted with the application it registered which button that was pressed and saved it together with a timestamp. Equally, every time the instruction program showed a new symbol it logged the symbol together with a timestamp. As a result we received two data sets of symbols with corresponding timestamps for each test person, resulting in a total of 200 records. Although the data was hard to interpret due to its compact look, a python script was written to structure the resulting timestamps into a more readable format. We calculated each RT by subtracting the timestamp when the instruction was given from the registered user input timestamp. Out of this we could calculate the mean RT for both groups LCC and HCC. The calculation of RT did not take in consideration if user input was a match with the instruction or not. It was still considered as an reaction.

@@

\section{Result}
From our results we can not draw any statistically significant differences in reaction time between the two groups high color contrast and low color contrast. This is most probably due to corrupt test data, a result from limitations in web application technology, see Section \ref{sec:discussion}. With user input errors we found that the low color contrast group produced significantly more user input errors than the hight color contrast group.

\subsection{Reaction Times (RT)}
Table~\ref{tab:groupRT} shows our average reaction time and standard deviation for each group. According to the results people in the high color contrast group had insignificant slower reaction times compared to the ones in the low color contrast group.

@@

To validate that the groups RTs are insignificantly different we assumed the two data sets where normal distributed based on the n>30 rule, and performed a two tailed t-test. The test produced a p-value of 0.17 which is bigger than a desired error margin ($\alpha$) of 5\%. The t-test is therefore proving that no differences in RT exists within our data. Regardless what the t-test says our standard deviations are indicating that something might be wrong with our data (see section \ref{sec:discussion}).

\subsection{Input Errors (IE)}
None of the groups where completely free from user input errors. As can be seen in Table~\ref{tab:userIE}, users that were given the LCC interface produced 475\% more UIE than the HCC group.

@@

\section{Discussion}\label{sec:discussion}
Our test results have shown that we can not increase the user reaction time by using a high color contrast interface when using a smart phone web based application. However we could see that the high color contrast interface was more reliable in performance when it comes to the amount of user input errors, according to our results.

What we have seen in our collected test data can strongly negotiate the legibility of the results. The calculated standard deviations gave us hints on that something was wrong. We dug deeper into each test persons results and found that the data must be corrupt. We found that we had randomly received negative RTs which in theory should be completely impossible. We were prepared that the timestamps could be off sync due to individual system clocks and no boot up pairing. However this would only have resulted in a consistent off sync. Instead we got inconsistent hick ups in our timestamps. We suspect this is a result from either hardware limitations or limitations in the JavaScript\footnote{A prototype based dynamic scripting language} engine for web applications on Android.

Our conclusion is that HCC interfaces are the preferred method to design interfaces as they deliver a more stable user interaction pattern. Apart from this, smart phone web based applications should only be used if sometimes slow technology, user input errors and long reaction times from the users can be accepted.

@@

\begin{description}
	\item[Small test group.] To be able to get a more statistical valid result the test group has to be bigger.
	\item[Many affecting parameters.] The prototype is simple in its appearance but there are still many parameters that can affect the users interaction and their RTs. Icons, icon size, optimal number of on screen buttons, size of hand-held device etc. All this, and more, should have been researched and taken into account if done again. 
	\item[Limited test environment.] Our tests were made using an Android based smart-phone\footnote{OnePlus Two \url{https://oneplus.net/2}} and the application were made using HTML5\footnote{HyperText Markup Language, markup standard for WWW}, CSS3\footnote{Cascading Style Sheet, styling standard for WWW} and JavaScript. We can not surely imply, without testing, that the same results would appear on an iPhone or any other smart-phone or even with a native application.
\end{description}

>>>>

\title{Can a High Color Contrast Touch Interface Increase User Reaction Time when Using a Smart Phone Web Based Application?}

@@

%%%%% BEGIN OF 2nd ADJUSTMENTS SECTION %%%%%
% Do NOT include an abstract in the outline and annotated bibliography
% (deliverable 2). For the full and final paper (deliverables 3 and 4a/b),
% you must uncomment the following two commands and place your text for
% the abstract in between them.
%
\begin{abstract}
This study aims to investigate if it is possible to receive better user performance when using a high-color-contrast compared to a low-color-contrast touch screen interface in a mobile web application. The performance was measured by combining the time each user took to respond to instructions (reaction time) and the number of input errors the user made. Received test data appeared to be corrupt, most probably due to limitations in web application technology, despite this our result show that the low color contrast interface produced 475\% more user input errors compared to the high color contrast interface. No difference in reaction time could be found between the two versions.
\end{abstract}
%
%
%This study aims to investigate if there is a performance difference between a high color contrast and a low color contrast touch screen interface. The performance was measured by combining the time each test subject took to respond on instructions (reaction time) and the number of input errors. The result shows that the high color contrast interface resulted in significantly fewer user input errors. There was no differences in reaction time between the two interfaces.
%%%%% END OF 2nd ADJUSTMENTS SECTION %%%%%

\section{Introduction}
The main goal with this paper is to evaluate if color contrast has a significant impact on usability in web based smart-phone applications. Consumer touch screen devices such as smart-phones have rapidly increased in number and availability recent years. The touch screen technology has made great advances~\cite{jennings2013touch} and it is frequently used as a way of receiving user input. When this technology moves to a broader audience, higher demands on usability needs to be set~\cite{gong2004guidelines}. To further explore how usability can be improved using touch screen technology we have in this paper investigated if reaction time (the time it takes for a user to react on instructions, further referred to as RT in this paper) and user input errors (the number of user input errors when interacting with a system, further referred to as UIE in this paper) can reach better results by increasing color contrast within the interface. With "better results" we mean fewer UIEs and faster RTs. Increased number of user input errors means that users repeatedly presses the wrong buttons which tells us that it is hard to accomplish desired tasks. Faster reaction times does not necessarily mean "better", but in this experiment we had tasks that should be performed as fast as possible which led us to the conclusion, faster reaction times are "better".

Contrast is determined by the difference in color and brightness of an object and objects within the same field of view. High contrast is the degree of difference between dark and light approaches the maximum possible difference. Low contrast is the opposite.

Earlier studies completed in this field have explored how we perceive different color contrast and how it can affect our reading performance~\cite{wu2003improving}. It has also been proved that chromaticity, contrast, and cone opponency in color space can affect RTs~\cite{mckeefry2003simple}. Many best practices for mobile development have also shown that "requirements for sufficient color contrast"~\cite{marcus2013design} must be met to best amplify the content. Still the question remains, can a high color contrast interface increase the the reaction time and decrease the amount of user input errors?

High RTs and low UIEs are especially desirable when designing user interfaces for situations with high demands on quick user input and low error tolerance such as in emergency situations. The results of studies like this are especially useful for designers and front end developers.

\section{Method}
In order to be able to test if color contrast has a substantial impact on the RT and UIE of a touch screen user interface we have designed a simple web application for an Android smart-phone. The application exists in two versions, one with low color contrast (referred to as LCC in this paper) and one with high color contrast (referred to as HCC in this paper) interface as can be seen in Fig~\ref{fig:application}. We conducted an A/B test to measure the performance difference between the versions.

@@

We designed the application by following design guidelines introduced and presented here~\cite{hoober2011designing,johnson2013designing,gong2004guidelines,norman2013design}. These guidelines can be seen as a set of rules to follow when designing any kind of user interface. We used guidelines to get proper sized buttons, good spacing between content and best positioning of navigation.

@@

The application consists of 4 buttons\footnote{On-screen button}, each button representing a function. To differentiate each button from the others we put a unique shape in each button. The shapes we used were a, square, circle, triangle and a rhomb (Fig~\ref{fig:application}).

To be able to give the test persons consistent instructions through the complete test we designed a program that presented the instructions on a secondary screen. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{instructions}
	\caption{Our program designed to present user instructions. Instructions were given in a series of 20. Between each instruction there was a delay set to be between 2.5 and 4 seconds. Each time a new symbol where showed a timestamp were logged in a unique user instruction file.
	\label{fig:instructions}}
\end{figure}
The instructions were a series of 20 instructions with shapes identical to the ones in the application buttons. The test persons were told to press the representative shape in the application as fast as they could in order to measure the reaction time. If they realized that they pressed the wrong button they were just told to continue the test by pressing the right button as fast as they realized. the application registered all wrong, non-matching input as an UIE. Each test was performed indoors with varying surroundings such as different lightning conditions and noise levels.

@@

\subsection{Evaluation of the A/B test}
Our data received from the tests consisted of UNIX timestamps\footnote{More info at: \url{http://www.unixtimestamp.com/}} measured in milliseconds. Each time the user interacted with the application it registered which button that was pressed and saved it together with a timestamp. Equally, every time the instruction program showed a new symbol it logged the symbol together with a timestamp. As a result we received two data sets of symbols with corresponding timestamps for each test person, resulting in a total of 200 records. Although the data was hard to interpret due to its compact look, a python script was written to structure the resulting timestamps into a more readable format. We calculated each RT by subtracting the timestamp when the instruction was given from the registered user input timestamp. Out of this we could calculate the mean RT for both groups LCC and HCC. The calculation of RT did not take in consideration if user input was a match with the instruction or not. It was still considered as a reaction.

@@

\section{Result}
From our results we can not draw any statistically significant conclusions about differences in reaction time between the two groups high color contrast and low color contrast. This is most probably due to corrupt test data, a result from limitations in web application technology, see Section \ref{sec:discussion}. With user input errors we found that the low color contrast group produced significantly more user input errors than the hight color contrast group.

\subsection{Reaction Times (RT)}
Table~\ref{tab:groupRT} shows the average reaction time and standard deviation for each group. According to the results people in the high color contrast group had insignificant slower reaction times compared to the ones in the low color contrast group.

@@

To check if the RTs in the two groups are significantly different we assumed the two data sets where normal distributed based on the n>30 preference, and performed a two tailed t-test. The test produced a p-value of 0.17 which is bigger than a desired error margin ($\alpha$) of 5\%. This shows that the difference in RT between the two groups is not statistically significant. Regardless what the t-test says our standard deviations are indicating that something might be wrong with our data (see section \ref{sec:discussion}).

\subsection{Input Errors (IE)}\label{subsec:InputErrors}
None of the groups where completely free from user input errors. As can be seen in Table~\ref{tab:userIE}, users that were given the LCC interface produced 475\% more UIE than the HCC group.

@@

\section{Discussion}\label{sec:discussion}
From our test results we cannot draw any conclusions about differences in reaction time between the high color contrast interface and the low color contrast interface. However we could see that the high color contrast interface was more reliable in performance when it comes to the amount of user input errors, according to our results.

What we have seen in our collected test data can strongly negotiate the legibility of the results. The calculated standard deviations gave us hints on that something was wrong. We dug deeper into each test persons results and found that the data must be corrupt. We found that we had randomly received negative RTs which in theory should be completely impossible. We were prepared that the timestamps could be off sync due to individual system clocks and no boot up pairing. However this would only have resulted in a consistent off sync. Instead we got inconsistent hiccups in our timestamps. We suspect this is a result from either hardware limitations or limitations in the JavaScript\footnote{A prototype based dynamic scripting language} engine for web applications on Android.

Although we can not draw any conclusions out from the results we have learned that smart phone web based applications should only be used if sometimes slow technology, user input errors and long reaction times from the user can be accepted. The HCC interface could also be seen as a preferred design method, even if the data was corrupt, because the users performed more consistent with UIEs compared to the users in the LCC group, see number of UIEs and their standard deviations in~\ref{subsec:InputErrors}.

@@

\begin{description}
	\item[Small test group.] To be able to get a more statistical valid result the test group has to be bigger. A good rule of thumb is at least 30.
	\item[Many affecting parameters.] The prototype is simple in its appearance but there are still many parameters that can affect the users interaction and their RTs. Icons, icon size, optimal number of on screen buttons, size of hand-held device etc. All this, and more, should have been researched and taken into account if done again. 
	\item[Limited test environment.] Our tests were made using an Android based smart-phone\footnote{OnePlus Two \url{https://oneplus.net/2}} and the application was made using HTML5\footnote{HyperText Markup Language, markup standard for WWW}, CSS3\footnote{Cascading Style Sheet, styling standard for WWW} and JavaScript. We can not surely imply, without testing, that we would get the same results on an iPhone or any other smart-phone, or even with a native application.
\end{description}
